Robust PCA and Higher Moments Distances for Botnet Detection

### 2019-04-21 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
copied from slides to paper

\usepackage{lineno,hyperref}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{algorithm,algorithmic}% http://ctan.org/pkg/algorithms
\usepackage{natbib}
\usepackage{supertabular}
\usepackage{fancybox}
\usepackage{acronym}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{setspace}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{courier}
\usepackage{amsfonts}
\usepackage{morefloats}
\usepackage{lipsum}
\usepackage{amsthm,amsmath}

% declarations
\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand*\justify{%
	\fontdimen2\font=0.4em% interword space
	\fontdimen3\font=0.2em% interword stretch
	\fontdimen4\font=0.1em% interword shrink
	\fontdimen7\font=0.1em% extra space
	%\hyphenchar\font=`\-% allowing hyphenation
}

@@

% Algorithmic modifications
\makeatletter
\newcommand{\ALOOP}[1]{\ALC@it\algorithmicloop\ #1%
  \begin{ALC@loop}}
\newcommand{\ENDALOOP}{\end{ALC@loop}\ALC@it\algorithmicendloop}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\BREAK}{\STATE \algorithmicbreak}
\makeatother

@@

@@

@@

...

@@

The proposed attack detection algorithm starts by the data pre-processing of a network traffic log containing IP, ports and timestamp of senders and receivers. During this step, the desired information is extracted in order to classify and count packets according to the origin and destination ports, and subsequently this information is grouped by minutes and by time frames.

With the data grouped into $Q$ time frames, the framework considers the time variations of the matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M\times{N}}$, with $q = 1, \ldots, Q$, in order to detect the attack.

According to flood and port scan attacks' behavior, flood attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of flood attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

Therefore, to detect flood attacks, it is necessary to calculate the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ of the zero mean samples given by

\begin{equation}\label{eq:eq02}
\boldsymbol{y}_{m}^{(q)} = \boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}.
\end{equation}

The set of obtained vectors $\boldsymbol{y}_{m}^{(q)}$ composes the zero mean matrix $\boldsymbol{Y}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{yy}^{(q)}$ can be calculated as follows

\begin{equation}\label{eq:eq03}
\boldsymbol{\hat{R}}_{yy}^{(q)} = \frac{1}{N}\boldsymbol{Y}^{(q)}\boldsymbol{Y}^{(q)^{\rm T}}.
\end{equation}

For the detection of the port scan attack, the main components are not dominated by the variables with large variance. Moreover, the portscan traffic presents a highly correlated network traffic. In order to exploit such structure, we compute the sample covariance $\boldsymbol{\hat{R}}_{zz}^{(q)}$ whose variables have zero mean and unitary standard deviation as follows

\begin{equation}\label{eq:eq04}
\boldsymbol{z}_{m}^{(q)} = \frac{\boldsymbol{x}_{m}^{(q)} - \bar{\boldsymbol{x}}_{m}^{(q)}}{\boldsymbol{\sigma}_{m}^{(q)}}.
\end{equation}

The set of vectors $\boldsymbol{z}_{m}^{(q)}$ composes the matrix $\boldsymbol{Z}^{(q)}$, then the sample covariance matrix $\boldsymbol{\hat{R}}_{zz}^{(q)}$ can be calculated via

\begin{equation}\label{eq:eq05}
\boldsymbol{\hat{R}}_{zz}^{(q)} = \frac{1}{N}\boldsymbol{Z}^{(q)}\boldsymbol{Z}^{(q)^{\rm T}}.
\end{equation}

Once the $\boldsymbol{\hat{R}}_{yy}^{(q)}$ and $\boldsymbol{\hat{R}}_{zz}^{(q)}$ have been obtained for flood and port scan attack detection, respectively, and since the next steps are the same for both sample covariance matrices, we refer to $\boldsymbol{\hat{R}}_{yy}$ and $\boldsymbol{\hat{R}}_{zz}$ as a matrix $\boldsymbol{C}$. Therefore, the following step of the algorithm is the eigenvalue decomposition (EVD), calculated according to (\ref{eq:eq06}), in order to obtain the vector of eigenvalues $\boldsymbol{e}^{(q)}$ associated with each matrix, according to (\ref{eq:eq060}).

\begin{equation}\label{eq:eq06}
\boldsymbol{C}^{(q)} = \boldsymbol{V}^{(q)}\boldsymbol{\Lambda}^{(q)}\boldsymbol{V}^{(q)^{\rm T}},
\end{equation}

\begin{equation}\label{eq:eq060}
\boldsymbol{e}^{(q)} = \rm diag(\boldsymbol{\Lambda}^{(q)}),
\end{equation}

where the operator diag$(\cdot)$ extracts the main diagonal of a matrix.

The eigenvalues should be sorted in descending order, i.e., $\lambda_{1}^{(q)} > \lambda_{2}^{(q)} > \lambda_{3}^{(q)} > ... > \lambda_{m}^{(q)}$. Therefore, the largest eigenvalue of the $q$-th time frame evaluated for the attack detect is given by $\lambda_{1}^{(q)}$.

>>>>

\usepackage{lineno,hyperref}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
\usepackage[linesnumbered, ruled, vlined]{algorithm2e}
\usepackage{natbib}
\usepackage{supertabular}
\usepackage{fancybox}
\usepackage{acronym}
\usepackage{array}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{tabularx}
\usepackage{multirow}
\usepackage{hhline}
\usepackage{setspace}
\usepackage{placeins}
\usepackage{longtable}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{courier}
\usepackage{amsfonts}
\usepackage{morefloats}
\usepackage{lipsum}
\usepackage{amsthm,amsmath}

@@

@@

Some techniques can make the data discriminative for better anomaly detection: Data engineering (i.e. sampling \cite{??}, aggregation \cite{acarali2016survey}); Feature extraction (i.e. decomposition \cite{??}, robust estimates \cite{??}).

Robust statistics produce methods that are not unduly affected by outliers.

Robust statistical methods have been developed for estimating location, scale, and regression parameters.

Robust estimates from contaminated data can denoise, remove anomalies or even separate them.

Anomaly detection based on robust estimates has been adopted for network anomaly detection, due to good results and its adaptability.

The distance between robust estimates and new observations can reveal anomalies.

Mahalanobis Distance (MD) is a generalized distance which is useful for determining the similarity between an unknown sample and a collection of known samples by, considering the correlations between the variables and their mean values.

\textbf{MD has been used for distance based anomaly detection with robust estimates} in many areas, and is commonly combined to Fast Minimum Covariance Determinant (Fast-MCD) algorithm \cite{rousseeuw1999fast} for robust estimates.

Network anomaly detection problems are usually characterized by skewed and imbalanced data \cite{Phua2004minority}.

Learning algorithms for imbalanced data has been a relevant research topic.

The fundamental issue with the imbalanced learning problem is the ability of imbalanced data to significantly compromise the performance of most standard learning algorithms \cite{he2008learning}.

Skewness and the kurtosis tests are very sensitive outlier detectors. The properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2008outlier}.

We propose a network anomaly detection approach based on robust moment estimates and distance analysis.

Considering that Fast-MCD is a widely adopted robust estimator of location and scatter for anomaly detection, we propose to extend Fast-MCD for computing robust skewness and kurtosis.

We also propose to use Mahalanobis Distance for detecting anomalies based on robust skewness and kurtosis, instead of the location-based usage of classical approaches.

Considering that it is hard to have labeled data in real problems \cite{osanaiye2016distributed}, such as for network attacks or frauds, also considering that known anomalies are constantly changing due to adversarial models or due to lack of regular patterns, we propose a Semi-Supervised approach based on robust moments for network anomaly detection.

@@

\cite{gu2007bothunter} (Bothunter: Detecting malware infection through ids-driven dialog correlation)

\cite{gu2008botminer} (Botminer: Clustering analysis of network traffic for protocol-and structure-independent botnet detection)

\cite{hubert2008outlier} (Outlier detection for skewed data)

\cite{garcia2014empirical} (An empirical comparison of botnet detection methods)

\cite{khattak2015botflex} (BotFlex: A community-driven tool for botnet detection)

\cite{wang2017botnet} (Botnet detection based on anomaly and community detection)

\cite{da2018online} (Online detection of Botnets on Network Flows using Stream Mining)

\cite{moustafa2019holistic} (A holistic review of Network Anomaly Detection Systems: A comprehensive survey). Network anomaly detection approaches are classified by \cite{moustafa2019holistic} in six categories: classification, clustering, deep learning, knowledge, combination and statistical-based.

\cite{Phua2004minority} (Minority Report in Fraud Detection: Classification of Skewed Data)

\cite{acarali2016survey} (Survey of approaches and features for the identification of HTTP-based botnet traffic)

\cite{lakhina2005mining} (Mining anomalies using traffic feature distributions)

\cite{goldberg2015importance} (The importance of features for statistical anomaly detection)

Clustering methods for anomaly detection rely on three key assumptions: legitimate data instances often fall into a cluster whereas attacks do not, legitimate data instances are usually located near the closest cluster centroid while anomaly ones are often far away from it, legitimate data instances fall into vast and dense clusters and anomalies into small or spare ones \cite{moustafa2019holistic}.

GMM is a statistical-based and parametric approach algorithm that estimates the distribution of the normal class from a training set and is typically based on a set of kernels \cite{moustafa2019holistic}.

The Mahalanobis Distance is a measure of the distance between a vector $\boldsymbol{x}$ and a distribution $\boldsymbol{X}$, introduced by P. C. Mahalanobis in 1936 \cite{mahalanobis1936generalized}. It is a multi-dimensional generalization for measuring how many standard deviations away $\boldsymbol{x}$ is from the mean $\boldsymbol{\bar{x}}$ and covariance $\boldsymbol{\hat{S}}$ of $\boldsymbol{X}$.

Mahalanobis Distance is defined as		
	\begin{equation}\label{eq:eq01}
		d(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\hat{S}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\hat{S}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}.
	\end{equation}
$\boldsymbol{x}$ is a vector of a new observation, $\bar{\boldsymbol{x}}$ is the mean vector, also referred as location, of known observations and $\boldsymbol{\hat{S}}$ is the covariance matrix, also referred as scatter, of known observations.

Classical estimates can be so strongly affected by contamination that diagnostic tools, such as the Mahalanobis Distances, become unable to detect the outliers. Therefore, we need reliable and robust estimators that can resist outliers contaminated data. Minimum Covariance Determinant (MCD) \cite{rousseeuw1984least} is a robust estimator commonly used for computing Mahalanobis Distance in order to detect anomalies.

\begin{figure}[h!]
     \centering
     \includegraphics[width=11cm]{figures/mahalanobis_robust.png}
     \caption{Mahalanobis Distance for non-robust and robust estimates.}
     \label{fig:fig04}
\end{figure}

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/mahalanobis_distances01.png}
     \caption{Mahalanobis Distance for non-robust and robust estimates.}
     \label{fig:fig05}
\end{figure}

The MCD estimator is one of the first affine equivariant and highly robust estimators of multivariate location and scatter.

Affine equivariant means that when the data are translated or subjected to a linear transformation, the resulting location and scatter will transform accordingly \cite{rousseeuw1984least,rousseeuw1999fast}.

Its objective is to find $h$ observations (out of $n$) whose covariance matrix has the lowest determinant.

The lowest determinant means the lowest distance generalized variance and high distance between the largest eigenvalue and the remaining [ref?].

MCD depends on random sampling, therefore it is not deterministic.

MCD has been applied in numerous fields (medicine, finance, image analysis, and chemistry) and been used to develop many robust multivariate techniques (such as Robust Principal Component Analysis (RPCA), factor analysis, and multiple regression).

The classical MCD has rarely been applied because it is hard to compute, since it requires the evaluation of all $\binom{n}{h}$ subsets of size $h$. % binomial of n over p

Its main use has started since the construction of the computationally efficient Fast-MCD algorithm \cite{rousseeuw1999fast}.

Fast-MCD is an iterative, approximate and resampling algorithm for the MCD.

It starts by random initial subset of size $p+1$ and performs concentration steps (C-steps) yielding consecutive h-subsets with decreasing covariance matrix determinant.

Only two C-steps are applied to each initial subset, and the 10 results with lowest determinant are kept.

C-steps are carried out until convergence and the best solution is kept.

Fast-MCD also is affine equivariant and not deterministic.

\textbf{Theorem 1: C-Step.}

Consider a data set $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ where $\boldsymbol{X} = \{\boldsymbol{x}_1,...,\boldsymbol{x}_n\}$ of $p$-variate observations. Let $\boldsymbol{h} \subset \{1,...,n\}$ with $|\boldsymbol{h}| = h$, where $\boldsymbol{h}$ is a subset of $h$ indexes from $n$ observations.

Let the mean $\bar{\boldsymbol{x}} \in \mathbb{R}^{1 \times p}$ be

\begin{equation}\label{eq:eq02}
	\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{j\in \boldsymbol{h}} \boldsymbol{x}_j, 
\end{equation}

the covariance matrix $\boldsymbol{\hat{S}} \in \mathbb{R}^{p \times p}$ be

\begin{equation}\label{eq:eq03}
	\boldsymbol{\hat{S}} = \displaystyle\frac{1}{h}\displaystyle\sum_{j\in \boldsymbol{h}} (\boldsymbol{x}_j - \bar{\boldsymbol{x}})(\boldsymbol{x}_j - \bar{\boldsymbol{x}})^\prime,
\end{equation}

and the relative distance be

\begin{equation}\label{eq:eq04}
	\boldsymbol{d}(i) = \sqrt{(\boldsymbol{x}_i - \bar{\boldsymbol{x}}) \boldsymbol{\hat{S}}^{-1}(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime}, 
\end{equation}

for $i = 1,...,n$.

Let $\boldsymbol{h}_1, \bar{\boldsymbol{x}}_1$ and $\boldsymbol{\hat{S}}_1$ and be the selected observations, the location and scatter of the initial subset, and $\boldsymbol{d}_1(i)$ be the initial distances for $i = 1,...,n$.

Now take $\boldsymbol{h}_2$ containing the indexes of lowest distances, such that $\{\boldsymbol{d}_1(i); i \in \boldsymbol{h}_2\} = \{(\boldsymbol{d}_1)_{1:n},...,(\boldsymbol{d}_1)_{h:n}\}$, where $(\boldsymbol{d}_1)_{1:n} \leq (\boldsymbol{d}_1)_{2:n} \leq ... \leq (\boldsymbol{d}_1)_{n:n}$ are the ordered distances, and compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$.

Then, $det(\boldsymbol{\hat{S}}_2) \leq det(\boldsymbol{\hat{S}}_1)$ with equality if and only if $\boldsymbol{\bar{x}}_2 = \boldsymbol{\bar{x}}_1$ and $\boldsymbol{\bar{S}}_2 = \boldsymbol{\hat{S}}_1$

The proof is given by Rousseeuw and Driessen \cite{rousseeuw1999fast}.

\textbf{The C-step can be algorithmically described as follows:}
\begin{algorithm}[H]\label{alg:alg01}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
	Given the initial $h-$subset $\boldsymbol{h}_1$ of indexes or the pair $(\boldsymbol{\bar{x}}_1, \boldsymbol{\hat{S}}_1)$\;
	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
		\If{$\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ exist}{
			Put $\boldsymbol{\bar{x}}_1 = \boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_1 = \boldsymbol{\hat{S}}_2$\;
		}
		Compute the distances $\boldsymbol{d}_1( i )$ for $i = 1,..., n$\;
		Sort $\boldsymbol{d}_1( i )$, which yields a index $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_1(\pi(n))$\;
		Put $\boldsymbol{h}_2 = \{ \pi(1), \pi(2), ..., \pi(h)\}$\;
		Compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$\;
		Compute $det(\boldsymbol{\hat{S}}_2)$ and $det(\boldsymbol{\hat{S}}_1)$\;			
	}
	\caption{C-Step}
\end{algorithm}

\textbf{Theorem 1} thus provides a partial idea for an algorithm:

\textit{Take many initial choices of $\boldsymbol{h}_1$ and apply C-steps to each until convergence (don't have determinant decreasing), and keep the result with lowest determinant}.

\textbf{Initial subset can be algorithmically described as:}
\begin{algorithm}[H]\label{alg:alg02}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{h}_1$}
	Given a data set $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ with $p$-variate observations\;
	Draw a random $(p + 1)$-subset $\boldsymbol{L}$\;
	Compute $\boldsymbol{\bar{x}}_1$ and $\boldsymbol{\hat{S}}_1$ of $\boldsymbol{L}$ and $det(\boldsymbol{\hat{S}}_1)$\;
	\If{$det(\boldsymbol{\hat{S}}_1) \leq 0$}{
		Add another random observation into $\boldsymbol{L}$\;
		Compute $\boldsymbol{\bar{x}}_1$ and $\boldsymbol{\hat{S}}_1$ of $\boldsymbol{L}$ and $det(\boldsymbol{\hat{S}}_1)$\;
	}
	Compute $\boldsymbol{d}_1^2(i) = (\boldsymbol{x}_i - \boldsymbol{\bar{x}}_1)^\prime \boldsymbol{\hat{S}}_1^{-1}(\boldsymbol{X}_i - \boldsymbol{\bar{x}}_0)$ for $i = 1, ..., n$\;
	Sort them into $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_0(\pi(n))$\;
	Put $\boldsymbol{h}_1 = \{\pi(1), ..., \pi(h)\}$\;
	\caption{Constructing the initial subset}
\end{algorithm}

\begin{algorithm}[H]\label{alg:alg03}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
	Given $\boldsymbol{X} \in \mathbb{R}^{n \times p}$, $h < n$, $p \geq 2$ and $n \leq 600$\;
	\For{500 times}{
		Construct initial $h$-subset $\boldsymbol{h}_1$ (Algorithm \ref{alg:alg02})\;
		Carry out two C-steps (Algorithm \ref{alg:alg01}) resulting $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$\;
	}		
	Put $\boldsymbol{h}_1$ as the 10 lowest $det(\boldsymbol{\hat{S}_2})$ computed above\;
	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
		Carry out C-steps (Algorithm \ref{alg:alg01})\;
	}
	\caption{Fast-MCD when $n \leq 600$}
\end{algorithm}

\begin{algorithm}[H]\label{alg:alg04}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
	Given $\boldsymbol{X} \in \mathbb{R}^{n \times p}$, $h < n$, $p \geq 2$ and $n > 600$\;
	Construct up to 5 disjoint random subsets of size $n_{sub} = 300$\;
	\For{each subset, repeat 100 times}{
		Construct initial $\boldsymbol{h}_1$ of size $h_{sub} = [n_{sub}(h/n)]$\;
		Carry out two C-steps of $n_{sub}$ and $h_{sub}$\;
		Keep the 10 best $(\boldsymbol{\bar{x}}_{sub}, \boldsymbol{\hat{S}}_{sub})$\;
	}		
	Pool the subsets, yielding the merged set of $n_{merged} = 1500$\;
	\For{50 best $(\boldsymbol{\bar{x}}_{sub}, \boldsymbol{\hat{S}}_{sub})$}{
		Carry out two C-steps of $n_{merged}$ and $h_{merged} = [n_{merged}(h/n)]$\;
		Keep the 10 best $(\boldsymbol{\bar{x}}_{merged}, \boldsymbol{\hat{S}}_{merged})$\;
	}
	\For{the full data set and $m_{full}$ best results}{
		Take several C-steps, using $n$ and $h$\;
		Keep the best final result $(\boldsymbol{\bar{x}}_{full}, \boldsymbol{\hat{S}}_{full})$\;
	}
	\caption{Fast-MCD when $n > 600$}
\end{algorithm}

@@

Evaluating network anomaly detection solutions remains a challenge due to the lack of up-to-date real-world data sets for training and testing \cite{osanaiye2016distributed}.

We propose to use CTU-13 \cite{garcia2014empirical} to evaluate our proposal, considering that the CTU-13 is a data set that aims to have a large capture of real botnet attacks and C\&C traffic mixed with normal and background traffic.

The CTU-13 data set consists in thirteen captures (called scenarios) of different botnet samples.

On each scenario it was executed a specific malware, which used several protocols and performed different actions.

The botnet traffic is divided into attacks and Command and Control (C\&C) traffic.

\textbf{Attacks:} Click Fraud, Port Scan, FastFlux, Compiled and Controled by Authors, SPAM and DDOS.

\textbf{C\&C:} IRC, P2P and HTTP.

The data set contains the following features:

Start Time, Duration, Protocol, Source Address, Source Port, Direction, Destination Address, Destination Port, State, Type of service from source to destination, Type of service from destination to source, Total of Packets, Total of bytes, Total bytes from source to destination, Label as normal or anomalous

\begin{table}[h!]
	\tiny
	\caption{CTU-13 Dataset Description}
	\label{tab:tab01}
	\begin{tabular}{| l | l | l | l | l | l | l | l | l | l | l | }
		\hline \rowcolor{Gray} \begin{tabular}[x]{@{}l@{}}Scenario\end{tabular}	& \begin{tabular}[x]{@{}l@{}}Malware\end{tabular}	 & \begin{tabular}[x]{@{}l@{}}Type\end{tabular}	& \begin{tabular}[x]{@{}l@{}}Total Flows\end{tabular} & \begin{tabular}[x]{@{}l@{}}Malicious\end{tabular} & \begin{tabular}[x]{@{}l@{}}C\&C\end{tabular} & \begin{tabular}[x]{@{}l@{}}Botnet\end{tabular}\\ \hline
			10 & neris & IRC, Spam, CF & 2,824,636 & 40,961 (1.45\%) & 341 (0.01\%) & 40,620 (1.44\%) \\ \hline
			11 & neris & IRC, Spam, CF & 1,808,122 & 20,941 (1.16\%) & 673 (0.04\%) & 20,268 (1.12\%) \\ \hline
			12 & rbot & IRC, PS, US & 4,710,638 & 26,822 (0.57\%) & 63 (0.00\%) & 26,759 (0.57\%) \\ \hline
			15 & rbot & IRC, DDoS, US & 1,121,076 & 2,580 (0.23\%) & 52 (0.00\%) & 2,528 (0.23\%) \\ \hline
			15-2 & virut & Spam, PS,HTTP & 129,832 & 901 (0.69\%) & 24 (0.02\%) & 877 (0.68\%) \\ \hline
			16 & menti & PS & 558,919 & 4,630 (0.83\%) & 199 (0.04\%) & 4,431 (0.79\%) \\ \hline
			16-2 & sogou & HTTP & 114,077 & 63 (0.06\%) & 26 (0.02\%) & 37 (0.03\%) \\ \hline
			16-3 & murlo & PS & 2,954,230 & 6,127 (0.21\%) & 1,074 (0.04\%) & 5,053 (0.17\%) \\ \hline
			17 & neris & IRC, Spam, CF,PS & 2,087,508 & 184,987 (8.86\%) & 2,973 (0.14\%) & 182,014 (8.72\%) \\ \hline
			18 & rbot & IRC, DDoS, US & 1,309,791 & 106,352 (8.12\%) & 33 (0.00\%) & 106,319 (8.12\%) \\ \hline
			18-2 & rbot & IRC, DDoS, US & 107,251 & 8,164 (7.61\%) & 2 (0.00\%) & 8,162 (7.61\%) \\ \hline
			19 & nsys.ay & P2P & 325,471 & 2,168 (0.67\%) & 25 (0.01\%) & 2,143 (0.66\%) \\ \hline
			15-3 & virut & Spam, PS, HTTP & 1,925,149 & 40,003 (2.08\%) & 536 (0.03\%) & 39,467 (2.05\%) \\ \hline
	\end{tabular}
\end{table}

Table \ref{tab:tab01} presents the malwares used for botnet attacks, the attack and C\&C types, the total number of flows, the number of malicious flows, which includes the C\&C and botnet flows. Table \ref{tab:tab01} also shows the ratio of C\&C and botnet flows.

Table \ref{tab:tab01} shows that CTU-13 is very imbalanced data set, when comparing the amount of malicious to normal traffic.

Malicious observations are beteeen $0.06\%$ up to $8.86\%$ of all flows.

The scenarios have between $107,251$ and $4,710,638$ flows.

The CTU-13 data set is also very skewed. Figure \ref{fig:fig06} exemplifies the skewed distribution of normal and anomalous flows, showing the flow duration of scenario 12.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/raw_distplot_capture20110812_Dur.png}
     \caption{Distribution of flow duration for scenario 12.}
     \label{fig:fig06}
\end{figure}

Methods for imbalance learning has attracted growing attention from both academia and industry, due to the restrictions imposed by imbalanced data to machine learning algorithms.

He and Garcia \cite{he2008learning} present a classification of imbalanced learning methods as sampling, cost-sensitive learning, kernel-based, and active learning.

Data aggregation is a widely adoted approach for feature engineering \cite{chandrashekar2014survey,acarali2016survey}.

Lakhina et al. \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies.

To deal with the data imbalance of CTU-13 data set, we adopt a sampling method based on data aggragation \cite{acarali2016survey} for feature generation, in order to obtain less imbalanced and more discrimnative data.

Based on related works \cite{lakhina2005mining,chandrashekar2014survey,acarali2016survey} and experience \cite{vieira2017model, galibus2017offline}, we selected some features for improving the anomaly detection.
The features are generated by a data aggregation into a selected window size. We evalute our proposal for windows with 0.15, 0.25, 1 and 2 seconds, for each scenario of the CTU-13 data set.

Following (2 slides) are the selected features:
\textbf{normal-flow-count:} count of normal (known) flow.
\textbf{background-flow-count :} count of background (unknown) flows.
\textbf{avg-duration:} average duration time of flows.
\textbf{n-conn:} count of flows.
\textbf{n-icmp:} count of ICMP packets.
\textbf{n-tcp:} count of TCP packets.
\textbf{n-udp:} count of UDP packets.
\textbf{n-dports<1024:} count of destinations ports lower than 1024.
\textbf{n-sports<1024:} count of source ports lower than 1024.
\textbf{n-sports>1024:} count of source ports higher than 1024.

Following are the remaining selected features:
	\textbf{n-s-a-p-address:} count of class A source address.
	\textbf{n-d-a-p-address:} count of class A destination address.
	\textbf{n-s-b-p-address:} count of class B source address.
	\textbf{n-d-b-p-address:} count of class B destination address.
	\textbf{n-s-c-p-address:} count of class C source address.
	\textbf{n-d-c-p-address:} count of class C destination address.
	\textbf{n-s-na-p-address:} count of class D* source address.
	\textbf{n-d-na-p-address:} count of class D* destination address.

Figures \ref{fig:fig07} and \ref{fig:fig08} show the distribution of average flow duration, for scenario 12, aggregated by windows of 0.15 and 2 seconds.

In comparison to Figure \ref{fig:fig06}, that shows the distribution of flow duration for the same scenario, we can see that the feature based on the average is more discriminative.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_0_15s_12_avg_duration.png}
     \caption{Distribution of average flow duration for scenario 12, aggregated by windows of 0.15 seconds.}
     \label{fig:fig07}
\end{figure}

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_2s_12_avg_duration.png}
     \caption{Distribution of average flow duration for scenario 12, aggregated by windows of 2 seconds.}
     \label{fig:fig08}
\end{figure}

@@

We propose to extend Fast-MCD for computing robust moments (skewness and kurtosis), in order to improve the anomaly detection in a skewed and imbalanced data set of network traffic with normal and botnet traffic.

Moments are a set of statistical parameters to measure a distribution. The arithmetic mean is the first general moment, the second is the variance, while skewness (asymmetry) is the third moment and kurtosis (excess) is the fourth moment.

The general formula of the $r$-th moment can be expressed as:
\begin{equation}\label{eq:eq05}
	m_r = \displaystyle\frac{1}{n}\displaystyle\sum_{i = 1}^{n}( x_i - \bar{x})^r. 
\end{equation}

Therefore, the Skewness is calculated as
\begin{equation}\label{eq:eq06}
	\boldsymbol{\bar{s}} = \frac{m_3}{m_2^{\frac{3}{2}}},
\end{equation}

and Kurtosis as
\begin{equation}\label{eq:eq07}
	\boldsymbol{\bar{k}} = \frac{m_4}{m_2^2} 
\end{equation}

For extending Fast-MCD, We initially modify the C-step Algorithm \ref{alg:alg01} to calculate Skewness and Kurtosis after to minimize $det(\boldsymbol{\hat{S}})$, accordingly to Algorithm \ref{alg:alg05}.

\begin{algorithm}[H]\label{alg:alg05}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$, $\boldsymbol{\bar{s}}_2$ and $\boldsymbol{\bar{k}}_2$}
	Given the initial $h-$subset $\boldsymbol{h}_1$ of indexes or the pair $(\boldsymbol{\bar{x}}_1, \boldsymbol{\hat{S}}_1)$\;
	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
		\If{$\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ exist}{
			Put $\boldsymbol{\bar{x}}_1 = \boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_1 = \boldsymbol{\hat{S}}_2$\;
		}
		Compute the distances $\boldsymbol{d}_1( i )$ for $i = 1,..., n$\;
		Sort $\boldsymbol{d}_1( i )$, which yields a index $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_1(\pi(n))$\;
		Put $\boldsymbol{h}_2 = \{ \pi(1), \pi(2), ..., \pi(h)\}$\;
		Compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$\;
		Compute $det(\boldsymbol{\hat{S}}_2)$ and $det(\boldsymbol{\hat{S}}_1)$\;			
	}
	Compute $\boldsymbol{\bar{s}}_2$ and $\boldsymbol{\bar{k}}_2$ for last $\boldsymbol{h}_2$\;
	\caption{C-Step with higher moments}
\end{algorithm}

The Algorithm \ref{alg:alg06} describes the Fast-MCD with higher moments when $n \leq 600$, through the modification of the original Fast-MCD for computing Skewness and Kurtosis from the robust estimate.

\begin{algorithm}[H]\label{alg:alg06}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$, $\boldsymbol{\bar{s}}_2$ and $\boldsymbol{\bar{k}}_2$}
	Given $\boldsymbol{X} \in \mathbb{R}^{n \times p}$, $h < n$, $p \geq 2$ and $n \leq 600$\;
	\For{500 times}{
		Construct initial $h$-subset $\boldsymbol{h}_1$ (Algorithm \ref{alg:alg02})\;
		Carry out two C-steps with higher moments (Algorithm \ref{alg:alg05}) resulting $\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$, $\boldsymbol{\bar{s}}_2$ and $\boldsymbol{\bar{k}}_2$\;
	}		
	Put $\boldsymbol{h}_1$ as the 10 lowest $det(\boldsymbol{\hat{S}_2})$ computed above\;
	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
		Carry out C-steps with higher moments (Algorithm \ref{alg:alg05})\;
	}
	\caption{Fast-MCD with higher moments when $n \leq 600$}
\end{algorithm}
